Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-04-08 submission/manuscript.Rmd
Update data and render prior to submission

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.

@@

regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
scfa_types <- c("acetate", "propionate", "butyrate")

regression <- read_tsv("../data/rf/regression_data_pool.tsv",
												col_types=cols(class=col_character(),
													input=col_character(),
													.default=col_double()
												)
 											) %>%
							mutate(microbiome=str_replace(input, "fit_", ""),
										microbiome=str_replace(microbiome, "_fit", "")
								) %>%
							select(class, microbiome, test_Rsquared) %>%
							filter(class %in% scfa_types) %>%
							filter(microbiome %in% regression_features) %>%
							mutate(data_type = case_when(
					       microbiome %in% c("genus", "otu")  ~ "16S",
								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
							 )) %>%
							group_by(class, microbiome, data_type) %>%
							summarize(median_Rsquared = median(test_Rsquared)) %>%
							ungroup()

@@

convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ec = "KEGG", pc2pathways = "pathway")

>>>>

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. Such studies are plagued by difficulties insuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation. Together, these findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.

@@

regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ko")
scfa_types <- c("acetate", "propionate", "butyrate")

regression <- read_tsv("../data/rf/regression_data_pool.tsv",
												col_types=cols(class=col_character(),
													input=col_character(),
													.default=col_double()
												)
 											) %>%
							mutate(microbiome=str_replace(input, "fit_", ""),
										microbiome=str_replace(microbiome, "_fit", "")
								) %>%
							select(class, microbiome, test_Rsquared) %>%
							filter(class %in% scfa_types) %>%
							filter(microbiome %in% regression_features) %>%
							mutate(data_type = case_when(
					       microbiome %in% c("genus", "otu")  ~ "16S",
								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
								 microbiome %in% c("pc2ko", "pc2pathways")  ~ "picrust"
							 )) %>%
							group_by(class, microbiome, data_type) %>%
							summarize(median_Rsquared = median(test_Rsquared)) %>%
							ungroup()

@@

convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ko = "KEGG", pc2pathways = "pathway")

### 2019-04-08 submission/manuscript.tex
Update data and render prior to submission

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
have anti-inflammatory and anti-proliferative activities (11). Direct
supplementation of SCFAs or feeding of fiber caused an overall reduction
in tumor burden in mouse models of colorectal cancer (12). These results
suggest that supplementation with fiber, which many colonic bacteria
ferment to produce SCFAs may confer beneficial effects against
colorectal cancer. Regardless, there is a lack of evidence that
increasing SCFA concentrations can protect against colorectal cancer in
humans. Case-control studies that have investigated possible
associations between SCFAs and colon tumor status have been plagued by
relatively small numbers of subjects, but have reported increased total
and relative fecal acetate levels and decreased relative fecal butyrate
concentrations in subjects with colonic lesions (13). In randomized
controlled trials fiber supplementation has been inconsistently
associated with protection against tumor formation and recurrence (14,
15). These findings temper enthusiasm for treatments that target the
production of SCFAs or for using them as biomarkers for protection
against tumorigenesis.

@@

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units (OTUs) based on 97\% similarity or into genera enabled
us to classify individuals as having adenomas or carcinomas using random
forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively, indicating that SCFAs had poor predictive power
on their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a subject's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained random forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14, which was observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained random forest regression models using metagenomic
sequence data to predict the concentration of the SCFAs (Figure S2).
Similar to the analysis using 16S rRNA gene sequence data, the
metagenomic data was not predictive of SCFA concentration. The largest
R\textsuperscript{2} was 0.055, which was observed when using KEGG data
to predict propionate concentrations (Figure 2B). Because of the limited
number of samples that we were able to generate metagenomic sequence
data from, we used our 16S rRNA gene sequence data to impute metagenomes
that were binned into metabolic pathways or KEGG categories using
PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
the imputed metagenomic data. The largest R\textsuperscript{2} was
0.097, which was observed when using KEGG data to predict butyrate
concentrations (Figure 2B). The inability to model SCFA concentrations
from microbiome data indicates that the knowledge of the abundance of
organisms and their genes was insufficient to predict SCFA
concentrations.

>>>>

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
have anti-inflammatory and anti-proliferative activities (11). Direct
supplementation of SCFAs or feeding of fiber caused an overall reduction
in tumor burden in mouse models of colorectal cancer (12). These results
suggest that supplementation with fiber, which many colonic bacteria
ferment to produce SCFAs may confer beneficial effects against
colorectal cancer. Regardless, there is a lack of evidence that
increasing SCFA concentrations can protect against colorectal cancer in
humans. Case-control studies that have investigated possible
associations between SCFAs and colon tumor status have been plagued by
relatively small numbers of subjects, but have reported increased total
and relative fecal acetate levels and decreased relative fecal butyrate
concentrations in subjects with colonic lesions (13). In randomized
controlled trials fiber supplementation has been inconsistently
associated with protection against tumor formation and recurrence (14,
15). Such studies are plagued by difficulties insuring subjects took the
proper dose and using subjects with prior polyp history who may be
beyond a point of benefiting from fiber supplementation. Together, these
findings temper enthusiasm for treatments that target the production of
SCFAs or for using them as biomarkers for protection against
tumorigenesis.

@@

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units (OTUs) based on 97\% similarity or into genera enabled
us to classify individuals as having adenomas or carcinomas using random
forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively, indicating that SCFAs had poor predictive power
on their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.15; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a subject's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained random forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14, which was observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained random forest regression models using metagenomic
sequence data to predict the concentration of the SCFAs (Figure S2).
Similar to the analysis using 16S rRNA gene sequence data, the
metagenomic data was not predictive of SCFA concentration. The largest
R\textsuperscript{2} was 0.055, which was observed when using KEGG data
to predict propionate concentrations (Figure 2B). Because of the limited
number of samples that we were able to generate metagenomic sequence
data from, we used our 16S rRNA gene sequence data to impute metagenomes
that were binned into metabolic pathways or KEGG categories using
PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
the imputed metagenomic data. The largest R\textsuperscript{2} was
0.085, which was observed when using KEGG data to predict propionate
concentrations (Figure 2B). The inability to model SCFA concentrations
from microbiome data indicates that the knowledge of the abundance of
organisms and their genes was insufficient to predict SCFA
concentrations.
